---
title: "Statistical Inference Introduction"
author: "Hans Elliott"
format:
  html:
    embed-resources: true
toc: true
---

Example:  

We observe inter-arrival times at a bus stop.  
We can model them with an exponential distribution.  
Say $X \sim Exp(\lambda)$.  

However, we don't observe $X$, just whether or not the inter-arrival time is greater than $m$ minutes, for some fixed $m$.  
Thus we observe $Y = \text{Indicator}(X > m)$, which we can model as $Y \sim Ber(p)$.  

$P(X > m) = \int_{m}^\infty \lambda e^{-\lambda x} dx = e^{-m \lambda}$  
$Y \sim Ber(e^{-m \lambda})$  

We can then deduce $\lambda$ from $p$ with some algebra:  
$$
p = e^{-m \lambda} \\
\lambda = \frac{\log p}{-m}
$$

(of course, we know $m$ since we are the ones waiting - it is not random).

An exponential random variable has mean $1/ \lambda$, so once we have an estimate of $\lambda$ we have an estimate for the average inter-arrival time even though we've only observed whether or not the time is greater than $m$ minutes. Cool!


```{r}
set.seed(1614)
#
# inter-arrival times
#
m <- 5
lambda <- 0.5
X <- rexp(10000, rate = lambda)
# population p for Y ~ Ber(p)
(p_true <- exp(-m * lambda))

#
# what we observe
#
Y <- (X > m)

(p_hat_mle <- mean(Y))

(lambda_hat <- log(p_hat_mle) / -m)

(avg_inter_arrival <- 1 / lambda_hat)
```

The error between the censored estimate of lambda and the true lambda is small relative to lambda.

```{r}
sim_lambda_error <- function(m, lambda, nsim = 10000) {
  X <- rexp(nsim, rate = lambda)
  lambda_hats <- vapply(m,
                        \(mi) log(mean(X > mi)) / -mi,
                        numeric(1))

  return(lambda_hats - lambda)
}

ev = 70
m <- seq(1, ev*6, by = 1)
plot(m, sim_lambda_error(m, 1/ev),
     main = "difference between censored est. of lambda and truth",
     sub = paste0("lambda = ", round(1/ev, digits = 4)),
     xlab = "m",
     ylab = "error"); abline(h = 0)

```


